---
title: "Analyze Cell-Free Synthesis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AnalyzeCellFreeSynthesis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(AnalyzeAssay)
```
This package is designed to aid in analyzing assay data from well plate experiments. It specifically focuses on analyzing kinetic cell-free protein synthesis data and this will be the focus of this vignette. The basic functions this package performs are:

* Averaging replicates
* Subtracting a background or baseline signal
* Normalizing all signals to a chosen signal

This package requires the tidyverse environment. Begin by installing or loading tidyverse
```{r}
library(tidyverse)
```

```{r, include = FALSE}
#set a theme for plots
ggplot2::theme_set(theme_minimal())
```

## Importing and formatting

Assay data is imported as two files. A file of data by well plate location and a file of identifying characteristics of the data (i.e. which proteins, DNAs, or Cell-free systems were used). Two sample data sets are available.

Assay data and identifiers are loaded into the workspace and merged with `import.assaydat`. This function takes the file path of the two input files. 

```{r}
system.file("extdata", "Data_1.xlsx", package = "AnalyzeAssay") -> PathtoDatInfile
system.file("extdata", "Data_1_Ids.xlsx", package = "AnalyzeAssay") -> PathtoIds
```

By default, the `Minutes` column of the assay data is not pivoted longer and data are join by `Well`. This function requires the `readxl` function to import xlsx notebooks. The returned file is kinetic data in a long format for analysis.

```{r}
library(readxl)
import.assaydat(PathtoDatInfile,
                PathtoIds
                ) -> AssayDat
head(AssayDat)
```

Assay statistics are then calculated with `dereplicate`. Mean, standard deviation, and the 95% confidence internval of the mean are calculated by default. GroupBy sets the variables names by which the statistics are calcualted; defaults are "DNA", "Protein" and "Minutes", others may include "Extract" or "CFS". The method `"MM"` can be called to calculate Max and Min instead. 
```{r}
#Rows containing replicates are collapsed and new columns are made for the calculated stats.

dereplicate(AssayDat,
            GroupBy = c("DNA", "Protein", "Minutes")
            ) -> AssayDat
head(AssayDat)
```

Dereplicated data can be directly plotted with `ggplot` which is included in the tidyverse environment. It may be illustrative to plot a subset of data, such as control proteins.
```{r, fig.width = 5}
#tidyverse filter a subset of the data
AssayDat %>% 
  filter(Protein %in% c("A", "B", "C", "D","E")) -> SubsetAssayDat

ggplot2::ggplot(SubsetAssayDat, 
       aes(x = Minutes,
           y = MeanSignal,
           color = Protein)
       ) + 
geom_point()
``` 

## Subtracting background signal

It can be useful to subtract a single signal from all out signals. In `background.signal`, the signal variable is selected with `BackgroundSignalType` and the chosen conditions is provided as quoted text to `NameofBackgroundSignal`. 

```{r, fig.width = 5}
subtract.background(SubsetAssayDat,
                    Signal = MeanSignal,
                    GroupBy = "Minutes",
                    BackgroundSignalType = Protein,
                    NameofBackgroundSignal = "B") -> BackgroundSubtracted_SubsetAssayDat

#Generates another column of BckSubSignal, which can be called in subsequent analyses or ggplot
head(BackgroundSubtracted_SubsetAssayDat)

ggplot2::ggplot(BackgroundSubtracted_SubsetAssayDat, 
       aes(x = Minutes,
           y = BckSubSignal,
           color = Protein)
       ) + 
geom_point()
```

## Normalization

Normalizing signal intensity can be helpful when comparing data sets.  

```{r, fig.width = 5}
#Generate two Assay Data subsets 
AssayDat %>% 
  filter(Protein %in% c("A", "B", "C", "D","E")) -> Subset1
AssayDat %>% 
  filter(Protein %in% c("AA", "AB", "AC", "AD", "AE")) -> Subset2

ggplot2::ggplot(Subset1, 
       aes(x = Minutes,
           y = MeanSignal,
           color = Protein)
       ) + 
geom_point()
ggplot2::ggplot(Subset2, 
       aes(x = Minutes,
           y = MeanSignal,
           color = Protein)
       ) + 
geom_point()
```


`normalize.bysignal` takes a signal type and the name of the signal and divides all the signals by the chosen signal. `GroupBy` can be used to normalize over a variables, Minutes is selected for kinetic data by default.
```{r, fig.width = 5}
#Combine the two data sets
AssayDat %>% 
  filter(Protein %in% c("A", "B", "C", "D","E", "AA", "AB", "AC", "AD", "AE")) -> SubsetAssayDat

#Normalize to the highest signal in the first plot.
normalize.bysignal(SubsetAssayDat,
                   Signal = "MeanSignal",
                   GroupBy = "Minutes",
                   NormalizationSignalType = Protein,
                   NameofNormalizationSignal = "C") -> Normalized_SubsetAssayDat

#Generates another column of NormalizedSignal, which can be called in subsequent analyses or ggplot
head(Normalized_SubsetAssayDat)

ggplot2::ggplot(Normalized_SubsetAssayDat, 
       aes(x = Minutes,
           y = NormalizedSignal,
           color = Protein)
       ) + 
geom_point()
```

